Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-08-08 docs/parasite_mortality_manuscript.tex
Added sentence on Adjei variability

The previous authors qualitatively detected PIHM
in 7 of the 10 datasets considered (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$). The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results showing that the Adjei Method has a high Type I error rate.

@@

Although superior to the Adjei Method, the Likelihood Method still cannot be applied to all real datasets. For host-parasite systems where host mortality occurs as a steep, non-linear function of parasite intensity only 80 hosts must be sampled to have an 80\% power in detecting PIHM. However, as the maximum slope of the survival function decreases and the function becomes somewhat linear, hundreds, or possibly thousands of hosts would have to be sampled to achieve the same result. This is consistent with previous studies which illustrate the difficulty of detecting PIHM from linear host survival functions \citep{Lanciani1989}.  While it may be feasible to sample several hundred invertebrates or small fish, even the smallest sample sizes are completely
unfeasible for many vertebrates, particularly the species of conservation
concern where addressing the impact of parasitism would be most important. An
even larger sample size would be required to identify PIHM when parasites are highly aggregated, mean infection intensity is
high, or parasite prevalence is low, all of which are
common in many parasitic helminths. Moreover, while linear functions make PIHM undetectable, at the other extreme, steep,
non-linear survival curves produce severely biased estimates of the survival
function. Give the interaction between all of these different factors, the
Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are
common and only moderately aggregated, and substantial host mortality occurs at relatively low parasite intensity.

>>>>

The previous authors qualitatively detected PIHM
in 7 of the 10 datasets considered (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the Likelihood Method did not detect a significant effect of PIHM was the Adjei dataset
for female \emph{S. tumbil}.  For this dataset there was a marginally significant effect
of PIHM ($\chi^2_{df=2} = 5.34; p = 0.069$). The Adjei Method detected PIHM in 9 of the 10 datasets (Table \ref{table:pihm}), consistent with our simulation results showing that the Adjei Method has a high Type I error rate.  Moreover, the Adjei Method estimates of the $LD_{50}$ were quite variable for the Crofton data, consistent with our simulation results that the Adjei Method $LD_{50}$ estimates could be imprecise for sample sizes of less than 1000 hosts (\emph{SI} Fig. 4-6).

@@

Although superior to the Adjei Method, the Likelihood Method still cannot be applied to all real datasets. For host-parasite systems where host mortality occurs as a steep, non-linear function of parasite intensity only 75 hosts must be sampled to have an 80\% power in detecting PIHM. However, as the maximum slope of the survival function decreases and the function becomes somewhat linear, hundreds, or possibly thousands of hosts would have to be sampled to achieve the same result. This is consistent with previous studies which illustrate the difficulty of detecting PIHM from linear host survival functions \citep{Lanciani1989}.  While it may be feasible to sample several hundred invertebrates or small fish, even the smallest sample sizes are completely
unfeasible for many vertebrates, particularly the species of conservation
concern where addressing the impact of parasitism would be most important. An
even larger sample size would be required to identify PIHM when parasites are highly aggregated, mean infection intensity is
high, or parasite prevalence is low, all of which are
common in many parasitic helminths. Moreover, while linear functions make PIHM undetectable, at the other extreme, steep,
non-linear survival curves produce severely biased estimates of the survival
function. Give the interaction between all of these different factors, the
Likelihood Method is probably limited to detecting PIHM in systems where greater than 100 hosts can be collected, parasites are
common and only moderately aggregated, and substantial host mortality occurs at relatively low parasite intensity.
